---
title: "WHO suicide dataset"
always_allow_html: true
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr, ggplot2)
library(broom)
library(knitr)
who_suicide_statistics <- read.csv(file = 'who_suicide_statistics.csv')
```

> *Warning*: this post talks about suicide and should be read with caution.

## The Data

[Exploratory data analysis](https://en.wikipedia.org/wiki/Exploratory_data_analysis) is essential to construct hypothesis. Today we'll explore the publicly available [WHO Suicide Statistics](https://www.kaggle.com/szamil/who-suicide-statistics) database (version from Kaggle). It consists of a single CSV table, with 43776 instances of merely 6 variables. We do not intend to speculate about suicide causes nor to make any judgements.  

```{r}
summary(who_suicide_statistics)
```

Clearly, we have a considerable amount of missing values, with data since 1979 to 2016, which is still quite recent. The sex and country variables must be converted to categorical ones:

```{r}
who_suicide_statistics$sex <- as.factor(who_suicide_statistics$sex)
who_suicide_statistics$country <- as.factor(who_suicide_statistics$country)
```

Next, the age variable should be an ordered factor:

```{r}
who_suicide_statistics$age <- factor(who_suicide_statistics$age, levels = c("5-14 years", "15-24 years", "25-34 years", "55-74 years", "75+ years"))
```

Let's take a look at our most important variable -- suicide number:

```{r, echo = FALSE, fig.height=3, fig.width=4}
hist(who_suicide_statistics$suicides_no, xlab = "Suicide Number", main = NA)
```

Clearly, the distribution is extremely skewed and zero-inflated, ranging from 0 to very high values. Let's create a proportional suicide number variable **(suicide_rate)**, defined by prop_suicide = suicides_no/population * 1000000 **(per million people)** and see its distribution:

```{r, fig.height=3, fig.width=4}
total_suicide_rate <- who_suicide_statistics %>% group_by(country, year) %>% summarise(rate_suicide = sum(suicides_no) * 1000000 / sum(population), .groups = "drop_last") %>% na.omit

hist(total_suicide_rate$rate_suicide, xlab = "Suicide rate", main = NA)
```

Much less variance, but still a very broad range. Let's summarise and plot some graphs to see the relationships between variables.

```{r}
library(ggplot2, dplyr)

total_suicide <- who_suicide_statistics %>% group_by(year, country) %>% summarise(total_suicide = sum(suicides_no, na.rm = T), .groups = "drop_last")
```
```{r, echo = FALSE, fig.height=5, fig.width=6}
ggplot(data = total_suicide, aes(year, total_suicide)) + geom_line(aes(color = country), na.rm = T, size = 1, alpha = 0.5, show.legend = F) + theme_bw() + labs(title ="Total suicide number per country over time") + ylab("Suicide number") + xlab("Year")

ggplot(data = total_suicide_rate, aes(year, rate_suicide)) + geom_line(aes(color = country), na.rm = T, size = 1, alpha = 0.2, show.legend = F) + theme_bw() + labs(title ="Suicide rate per country over time") + ylab("Suicide rate (per million people)") + xlab("Sex")
```
```{r, echo = FALSE, fig.width=4, fig.height=5}
who_suicide_statistics %>% group_by(year, country, sex) %>% summarise(rate_suicide = sum(suicides_no) * 1000000 / sum(population), .groups = "drop_last") %>% na.omit %>% ggplot(aes(sex, rate_suicide)) + geom_violin(aes(fill = sex), size = 1, alpha = 0.3, show.legend = F) + geom_boxplot(aes(fill = sex), width = 0.05, size = 0.5, outlier.shape = NA, show.legend = F) + theme_bw() + labs(title = "Suicide rate per sex") + ylab("Suicide rate (per million people)") + xlab("Sex")

```

Men have higher suicide rates overall Let's see which countries have the most and least suicides:

**Top 10 countries and correspondent years with highest suicide rates**
```{r, echo = FALSE}
head(total_suicide_rate %>% arrange(desc(rate_suicide)), n = 10) %>% kable()
```

**Top 10 countries and correspondent years with lowest positive suicide rates**
```{r, echo = FALSE}
head(total_suicide_rate %>% filter(rate_suicide > 0) %>% arrange(rate_suicide), n = 10) %>% kable()
```

Now let's take an average over the last five years of data and see again the highs and lows:

**Top 20 countries with highest suicide rates (2012-2016 average)**
```{r, echo = FALSE}
head(total_suicide_rate %>% filter(year >= 2012) %>% group_by(country) %>% summarise(rate_suicide = mean(rate_suicide, na.rm = T)) %>% arrange(desc(rate_suicide)), n = 20) %>% kable()
```

**Top 20 countries with lowest positive suicide rates (2012-2016 average)**
```{r, echo = FALSE}
head(total_suicide_rate %>% filter(year >= 2012) %>% group_by(country) %>% summarise(rate_suicide = mean(rate_suicide, na.rm = T)) %>% filter(rate_suicide > 0) %>% arrange(rate_suicide), n = 20) %>% kable()
```

Let's see if there's any relationship between suicide rates (2012-2016) and [Democracy Index](https://dataspace.princeton.edu/jspui/bitstream/88435/dsp017p88ck01w/1/EIU-Democracy-Index-2015.pdf) (2015) calculated by The Economist group. The democracy index data was manually curated to correspond to country names present in the WHO dataset.

```{r, fig.height=3, fig.width=4, warning=FALSE, message=FALSE}
democracy <- read.csv(file = "democracy_index_2015.csv")

democracy_compare_data <- total_suicide_rate %>% filter(year >= 2012) %>% filter(country %in% as.character(unique(democracy$Country))) %>% group_by(country) %>% summarise(rate_suicide = mean(rate_suicide, na.rm = T)) %>% arrange(country)

democracy <- democracy %>% filter(Country %in% as.character(unique(democracy_compare_data$country))) %>% arrange(Country)

democracy_compare_data$overall_score <- democracy$Overall_score

ggplot(data = democracy_compare_data, aes(overall_score, rate_suicide)) + geom_point(size = 2, alpha = 0.75, colour = "dark blue") + theme_bw() + geom_smooth(formula = y ~ x, method = "loess", se = F) + xlab("Democracy score (overall)") + ylab("Suicide rate (per million people)")

tidy(cor.test(democracy$Overall_score, democracy_compare_data$rate_suicide, method = "pearson")) %>% kable()
tidy(cor.test(democracy$Overall_score, democracy_compare_data$rate_suicide, method = "spearman")) %>% kable()
```

There's a **weak** (R = 0.307) but significant **positive Pearson correlation between the Democracy Index and suicide rates**. However, there are many confounding factors here, as more democratic countries are in general richer and may report suicide statistics with better accuracy. Also, there are huge cultural differences between countries. Among highly democratic nations the correlation is near zero:

```{r, fig.height=3, fig.width=4}
democracy_compare_data %>% filter(overall_score > 6) %>% ggplot(aes(overall_score, rate_suicide)) + geom_point(size = 2, alpha = 0.75, colour = "dark blue") + theme_bw() + geom_smooth(formula = y ~ x, method = "loess", se = F) + xlab("Democracy score (overall)") + ylab("Suicide rate (per million people)")
```


Gross domestic product based on purchasing-power-parity (PPP) *per capita* GDP values (2015) in international dollars were obtained from the [International Monetary Fund (IMF)](https://www.imf.org/external/pubs/ft/weo/2016/02/weodata/weorept.aspx?sy=2015&ey=2015&ssd=1&sort=country&ds=.&br=1&pr1.x=16&pr1.y=7&c=512%2C668%2C914%2C672%2C612%2C946%2C614%2C137%2C311%2C962%2C213%2C674%2C911%2C676%2C193%2C548%2C122%2C556%2C912%2C678%2C313%2C181%2C419%2C867%2C513%2C682%2C316%2C684%2C913%2C273%2C124%2C868%2C339%2C921%2C638%2C948%2C514%2C943%2C218%2C686%2C963%2C688%2C616%2C518%2C223%2C728%2C516%2C558%2C918%2C138%2C748%2C196%2C618%2C278%2C522%2C692%2C622%2C694%2C156%2C142%2C624%2C449%2C626%2C564%2C628%2C565%2C228%2C283%2C924%2C853%2C233%2C288%2C632%2C293%2C636%2C566%2C634%2C964%2C238%2C182%2C662%2C453%2C960%2C968%2C423%2C922%2C935%2C714%2C128%2C862%2C611%2C135%2C321%2C716%2C243%2C456%2C248%2C722%2C469%2C942%2C253%2C718%2C642%2C724%2C643%2C576%2C939%2C936%2C644%2C961%2C819%2C813%2C172%2C199%2C132%2C733%2C646%2C184%2C648%2C524%2C915%2C361%2C134%2C362%2C652%2C364%2C174%2C732%2C328%2C366%2C258%2C734%2C656%2C144%2C654%2C146%2C336%2C463%2C263%2C528%2C268%2C923%2C532%2C738%2C944%2C578%2C176%2C537%2C534%2C742%2C536%2C866%2C429%2C369%2C433%2C744%2C178%2C186%2C436%2C925%2C136%2C869%2C343%2C746%2C158%2C926%2C439%2C466%2C916%2C112%2C664%2C111%2C826%2C298%2C542%2C927%2C967%2C846%2C443%2C299%2C917%2C582%2C544%2C474%2C941%2C754%2C446%2C698%2C666&s=PPPPC&grp=0&a=).

```{r, warning=FALSE}
gdppc <- read.csv("WEO_Data.xls", sep = "\t")
gdppc$X2015 <- as.numeric(as.character(gdppc$X2015))

gdp_compare_data <- total_suicide_rate %>% filter(year >= 2012) %>% filter(country %in% as.character(unique(gdppc$Country))) %>% group_by(country) %>% summarise(rate_suicide = mean(rate_suicide, na.rm = T)) %>% arrange(country)

gdppc <- gdppc %>% filter(Country %in% as.character(unique(gdp_compare_data$country))) %>% arrange(Country)
```
```{r, echo = FALSE, fig.height=3}
hist(gdppc$X2015, xlab = "PPP GDP per capita (2015)", main = NA)
hist(log(gdppc$X2015), xlab = "Log of PPP GDP per capita (2015)", main = NA)
```

As the GDP variable is heavily skewed, it's better to visualize it using its log transform:

```{r, echo = FALSE, fig.height=3, fig.width=4}
ggplot(data = gdp_compare_data, aes(log(gdppc$X2015), rate_suicide)) + geom_point(size = 2, alpha = 0.75, colour = "dark blue") + geom_smooth(formula = y~x, method = "loess", se=F) + theme_bw() + xlab("PPP GDP per capita (log)") + ylab("Suicide rate (per million people)")
```
```{r}
tidy(cor.test(gdppc$X2015, gdp_compare_data$rate_suicide, method = "spearman")) %>% kable()
```

There does not seem to exist an apparent association between suicide rates and *per capita* GDP income.

## Gender Ratios

```{r fig.height=3}
female_rates <- who_suicide_statistics %>% filter(year >= 2012) %>% group_by(country, sex) %>% summarise(rate_suicide = sum(suicides_no) * 1000000 / sum(population), .groups = "drop_last") %>% na.omit %>% arrange(country) %>% filter(sex == "female")

male_rates <- who_suicide_statistics %>% filter(year >= 2012) %>% group_by(country, sex) %>% summarise(rate_suicide = sum(suicides_no) * 1000000 / sum(population), .groups = "drop_last") %>% na.omit %>% arrange(country) %>% filter(sex == "male")

gender_ratio <- data.frame(country = female_rates$country, ratio = male_rates$rate_suicide / female_rates$rate_suicide) %>% na.omit() %>% filter(is.finite(ratio))

hist(gender_ratio$ratio, main = NA, xlab = "Gender Ratio")

gender_ratio_gdp <- gender_ratio %>% filter(country %in% as.character(unique(gdppc$Country)))
gdppc_gender <- gdppc %>% filter(Country %in% as.character(unique(gender_ratio_gdp$country)))
#ggplot(data = gender_ratio_gdp, aes(log(gdppc_gender$X2015), ratio)) + geom_point(size = 2, alpha = 0.75, colour = "dark blue") + theme_bw() + geom_smooth(se=F)

tidy(cor.test(gender_ratio_gdp$ratio, gdppc_gender$X2015)) %>% kable()

gender_ratio_dem <- gender_ratio %>% filter(country %in% as.character(unique(democracy$Country)))
democracy_gender <- democracy %>% filter(Country %in% as.character(unique(gender_ratio_dem$country)))
#ggplot(data = gender_ratio_dem, aes(democracy_gender$Overall_score, ratio)) + geom_point(size = 2, alpha = 0.75, colour = "dark blue") + theme_bw() + geom_smooth(se=F)

tidy(cor.test(gender_ratio_dem$ratio, democracy_gender$Overall_score)) %>% kable()

```

There does not seem to be any association between gender ratios and Democracy Index nor *per capita*
GDP.

**Top 10 countries with highest gender ratios (male-to-female) 2012-2016**
```{r}
head(gender_ratio %>% arrange(desc(ratio)), n = 20) %>% kable()
```

**Top 10 countries with lowest positive gender ratios (male-to-female) 2012-2016**
```{r}
head(gender_ratio %>% filter(ratio > 0) %>% arrange(ratio), n = 20) %>% kable()
```

## Age

Elderly suicide is an increasingly troublesome concern as the population grows older.

```{r}
elderly_data <- who_suicide_statistics %>% filter(year >= 2012) %>% filter(age == "55-74 years" | age == "75+ years") %>% group_by(country) %>% summarise(rate_suicide = sum(suicides_no) * 1000000 / sum(population)) %>% na.omit %>% arrange(desc(rate_suicide))
```

**Top 10 countries with highest elderly suicide rates (2012-2016)**
```{r}
head(elderly_data, n = 10) %>% kable()
```

This, however, can be biased due to a higher overall higher incidence of suicides in some countries. Thus, let's calculate the percentage of total suicides that are elderly ones (55+ years).

```{r}
total_elderly <- who_suicide_statistics %>% filter(year >= 2012) %>% filter(age == "55-74 years" | age == "75+ years") %>% group_by(country) %>% summarise(total_suicide = sum(suicides_no)) %>% na.omit

total_2012_16 <- total_suicide %>% filter(year >= 2012) %>% group_by(country) %>% summarise(total_suicide = sum(total_suicide, na.rm = T)) %>% filter(country %in% as.character(unique(total_elderly$country)))

elderly_proportion <- data.frame(country = total_elderly$country, proportion = total_elderly$total_suicide / total_2012_16$total_suicide)

elderly_proportion <- elderly_proportion[is.finite(elderly_proportion$proportion), ]
```

**Top 10 countries with highest elderly suicide proportion (2012-2016)**
```{r}
head(elderly_proportion %>% arrange(desc(proportion)), n = 10) %>% kable()
```

## USA and Brazil: a case-study

I've selected two countries for further analysis: **Brazil and USA**, both very big countries with reliable data.

```{r}
BR_data <- subset(who_suicide_statistics, country == "Brazil")

US_data <- subset(who_suicide_statistics, country == "United States of America")
```
```{r, echo = FALSE, fig.height=4, fig.width=5}
ggplot(data = subset(total_suicide_rate, country == "Brazil"), aes(year, rate_suicide)) + geom_line(size = 1, color = "dark blue") + theme_bw() +labs(title = "Suicide rates in Brazil") + xlab("Year") + ylab("Suicide rate (per million people)")

ggplot(data = subset(total_suicide_rate, country == "United States of America"), aes(year, rate_suicide)) + geom_line(size = 1, color = "dark blue") + theme_bw() +labs(title = "Suicide rates in US") + xlab("Year") + ylab("Suicide rate (per million people)")
```

```{r, echo = FALSE, fig.height=5, fig.width=3}
BR_data %>% group_by(year, sex) %>% summarise(rate_suicide = sum(suicides_no) * 1000000 / sum(population), .groups = "drop_last") %>% na.omit %>% ggplot(aes(sex, rate_suicide)) + geom_violin(aes(fill = sex), size = 1, alpha = 0.3, show.legend = F) + geom_boxplot(aes(fill = sex), width = 0.05, size = 0.5, outlier.shape = NA, show.legend = F) + theme_bw() + labs(title = "Suicide rate per sex (BR)") + xlab("Sex") + ylab("Suicide rate (per million people)")

US_data %>% group_by(year, sex) %>% summarise(rate_suicide = sum(suicides_no) * 1000000 / sum(population), .groups = "drop_last") %>% na.omit %>% ggplot(aes(sex, rate_suicide)) + geom_violin(aes(fill = sex), size = 1, alpha = 0.3, show.legend = F) + geom_boxplot(aes(fill = sex), width = 0.05, size = 0.5, outlier.shape = NA, show.legend = F) + theme_bw() + labs(title = "Suicide rate per sex (US)") + xlab("Sex") + ylab("Suicide rate (per million people)")
```

Gender differences can be calculated over time:

```{r, echo = FALSE, fig.height=4, fig.width=5}
BR_data %>% group_by(year, sex) %>% summarise(rate_suicide = sum(suicides_no) * 1000000 / sum(population), .groups = "drop_last") %>% na.omit %>% ggplot(aes(year, rate_suicide)) + geom_line(aes(color = sex), size = 1) + theme_bw() + labs(title = "Suicide rate over time per sex (BR)") + xlab("Year") + ylab("Suicide rate (per million people)")

US_data %>% group_by(year, sex) %>% summarise(rate_suicide = sum(suicides_no) * 1000000 / sum(population), .groups = "drop_last") %>% na.omit %>% ggplot(aes(year, rate_suicide)) + geom_line(aes(color = sex), size = 1) + theme_bw() + labs(title = "Suicide rate over time per sex (US)") + xlab("Year") + ylab("Suicide rate (per million people)")
```
```{r}
sex_US_data <- US_data %>% group_by(year, sex) %>% summarise(rate_suicide = sum(suicides_no) * 1000000 / sum(population), .groups = "drop_last") %>% na.omit

sex_BR_data <- BR_data %>% group_by(year, sex) %>% summarise(rate_suicide = sum(suicides_no) * 1000000 / sum(population), .groups = "drop_last") %>% na.omit

US_data_sexratio <- data.frame(year = subset(sex_US_data, sex == "male")$year, ratio = subset(sex_US_data, sex == "male")$rate_suicide / subset(sex_US_data, sex == "female")$rate_suicide, country = "US")

BR_data_sexratio <- data.frame(year = subset(sex_BR_data, sex == "male")$year, ratio = subset(sex_BR_data, sex == "male")$rate_suicide / subset(sex_BR_data, sex == "female")$rate_suicide, country = "BR")

data_sexratio <- rbind(US_data_sexratio, BR_data_sexratio)
```
```{r, echo = FALSE, fig.height=4, fig.width=5}
ggplot(data = data_sexratio, aes(year, ratio)) + geom_line(aes(color = country), size = 1) + theme_bw() + labs(title = "Suicide rate men-to-women ratio") + xlab("Year") + ylab("Suicide rate men-to-women ratio")

```

In Brazil, **suicide rates for men have been steadily increasing since the 1980s**, while rates for women have stayed roughly the same. In the US, however, suicide rates for men increased during the 80s (not followed by an increase in women's rates), decline in the 2000s and has been increasing since 2005-6. This increase is now followed by a similar (but smaller) one in women's rates. Thus, the men-to-women ratio increased with time in Brazil and decreased only after 2000 in the US. In 2015, **for each woman**, **4-4.5 men** have ended their lives in Brazil or in the US.

```{r}
age_data_usbr <- who_suicide_statistics %>% group_by(year, country, age) %>% summarise(rate_suicide = sum(suicides_no) * 1000000 / sum(population), .groups = "drop_last") %>% na.omit
```
```{r, echo = FALSE, fig.height=5, fig.width=6}
ggplot(data = subset(age_data_usbr, country == "Brazil"), aes(year, rate_suicide)) + geom_line(aes(color = age), size = 1) + theme_bw() + labs(title = "Suicide rates per age group (BR)") + xlab("Year") + ylab("Suicide rate (per million people)")

ggplot(data = subset(age_data_usbr, country == "United States of America"), aes(year, rate_suicide)) + geom_line(aes(color = age), size = 1) + theme_bw() + labs(title = "Suicide rates per age group (US)") + xlab("Year") + ylab("Suicide rate (per million people)")

```

Both countries present highest suicide rates for the elderly. However, in both cases, **the gap between adults (25-34 years) and elderly (55+ years) is getting narrower** since the 2000s, which shows that adult suicide is more likely now than compared to the past (1990s).

```{r}
age_gender_usbr <- who_suicide_statistics %>% group_by(sex, year, country, age) %>% summarise(rate_suicide = sum(suicides_no) * 1000000 / sum(population), .groups = "drop_last") %>% na.omit
```
```{r, echo = FALSE, fig.height=5, fig.width=11}
ggplot(data = subset(age_gender_usbr, country == "Brazil"), aes(year, rate_suicide)) + geom_line(aes(color = age), size = 1) + theme_bw() + labs(title = "Suicide rates per age group and sex (BR)") + facet_grid(cols = vars(sex)) + xlab("Year") + ylab("Suicide rate (per million people)")

ggplot(data = subset(age_gender_usbr, country == "United States of America"), aes(year, rate_suicide)) + geom_line(aes(color = age), size = 1) + theme_bw() + labs(title = "Suicide rates per age group and sex (US)") + facet_grid(cols = vars(sex))+ xlab("Year") + ylab("Suicide rate (per million people)")
```

Interestingly, the high elderly suicide rate is apparently accounted for by only male people. **There's practically no age gap among women**. This suggests that elderly suicide is almost exclusively a male issue in these countries.

## Conclusion

This exploratory analysis is descriptive and serves the purpose to inform about overall characteristics and trends in global suicide reports provided by the WHO. Suicide is a complex social phenomenon and should not be interpreted simplistically. Still, the huge difference between genders in the age gap is of interest.
